---
title: "Practical Machine Learning - Project"
author: "karunesh bhardwaj"
date: "October 14, 2016"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

# 1. Overview
This paper is the final report of the project from Coursera's course Practical Machine Learning, which is part of Data Science specialization track. This report was created using RStudio, using markdown file function knitr and published in html format. The outcome of this analysis/study is to predict the course quiz questions. This is described by the variable "classe" in training set. The machine learning algorithm explained here is applied to the 20 test cases available in the test data and the predictions are submitted in appropriate format to the Course Project Prediction Quiz for automated grading.

# 2. Background
Using devices such as Jawbone Up, Nike FuelBand, and Fitbit it is now possible to collect a large amount of data about personal activity relatively inexpensively. These type of devices are part of the quantified self movement - a group of enthusiasts who take measurements about themselves regularly to improve their health, to find patterns in their behavior, or because they are tech geeks. One thing that people regularly do is quantify how much of a particular activity they do, but they rarely quantify how well they do it. In this project, your goal will be to use data from accelerometers on the belt, forearm, arm, and dumbell of 6 participants. They were asked to perform barbell lifts correctly and incorrectly in 5 different ways. More information is available from the website here: http://groupware.les.inf.puc-rio.br/har (see the section on the Weight Lifting Exercise Dataset). 

# 3. Data Loading and Exploratory Analysis

### i) Dataset Overview

The training data for this project are available here: 
https://d396qusza40orc.cloudfront.net/predmachlearn/pml-training.csv
The test data are available here:
https://d396qusza40orc.cloudfront.net/predmachlearn/pml-testing.csv

### ii) Prepare Environment

Populate and load all R libraries
```{r}
rm (list=ls())  # clean up memory
setwd("C:/Rproject/kb/MachineLearning")
library(knitr)
library(caret)
library(rpart)
library(rpart.plot)
library(rattle)
library(randomForest)
library(corrplot)
set.seed(12000)

```

### iii) Data Loading and Cleaning
```{r}
# Configure URL for data download

kbTrainUrl <- "https://d396qusza40orc.cloudfront.net/predmachlearn/pml-training.csv"
kbTestUrl <- "https://d396qusza40orc.cloudfront.net/predmachlearn/pml-testing.csv"

# Download the data

kbTraining <- read.csv(url(kbTrainUrl))
kbTesting <- read.csv(url(kbTestUrl))

# create a partition with the training dataset

kbinTraining <- createDataPartition(kbTraining$classe, p=0.7, list=FALSE)
kbTrainSet <- kbTraining[kbinTraining, ]
kbTestSet <- kbTraining[-kbinTraining, ]
dim(kbTrainSet)
dim(kbTestSet)

```
##### Both kbTrainSet and kbTestSet have 160 variables. Any NA in these variable can be cleaned by using following procedure. The Near Zero Variance (NZV) and ID variables are also removed. 

```{r}
# remove Near Zero Variance variables

NVZ <- nearZeroVar(kbTrainSet)
kbTrainSet <- kbTrainSet[, -NVZ]
kbTestSet <- kbTestSet[, -NVZ]
dim(kbTrainSet)
dim(kbTestSet)

# remove any NA variables

NAAny <- sapply(kbTrainSet, function(x) mean(is.na(x))) > 0.95
kbTrainSet <- kbTrainSet[, NAAny==FALSE]
kbTestSet <- kbTestSet[, NAAny==FALSE]
dim(kbTrainSet)
dim(kbTestSet)

# remove ID variables (col 1 to 5)
kbTrainSet <- kbTrainSet[, -(1:5)]
kbTestSet <- kbTestSet[, -(1:5)]
dim(kbTrainSet)
dim(kbTestSet)

```
#### By above cleaning process, the total number of variable for the analysis has been reduced to 54 variables.

### iv) Correlation Analysis
####Before proceeding to the modeling procedure, a correlation between variables must be analysed.

```{r}
kbCorMatrix <- cor(kbTrainSet[, -54])

corrplot(kbCorMatrix, order = "FPC", method = "color", type = "lower", tl.cex = 0.8, tl.col = rgb(0, 0, 0))

```

#### Dark color in the grap shows highly correlated variables, which are very few. Therefore, there is no need to preprocess the data with Principal Components Analysis (PCA).

# 4. Building Prediction Model

Three methods will be applied to model the regressions to the Training Dataset and the best one, which give higher accuracy will be used for the quiz predictions. The methods are: Random Forests, Decision Tree and Generalized Boosted Model.

### i) Random Forest Method

```{r}
# Fit Model

set.seed(12000)
RFcontrol <- trainControl(method="cv", number=3, verboseIter = FALSE)
FitRmodandForest <- train(classe ~ ., data=kbTrainSet, method="rf", trControl=RFcontrol)
FitRmodandForest$finalModel

# prediction on Test dataset

RandForestprediction <- predict(FitRmodandForest, newdata=kbTestSet)
RandForestConfMatrix <- confusionMatrix(RandForestprediction, kbTestSet$classe)
RandForestConfMatrix

# Plot Matrix Result

plot(RandForestConfMatrix$table, col=RandForestConfMatrix$byClass, main=paste("Random Forest - Accuracy =", round(RandForestConfMatrix$overall ["Accuracy"], 4)))

```

### ii) Decisioin Tree Method
```{r}
# Fit Model
set.seed(12000)
DecTreeFitModel <- rpart(classe ~ ., data = kbTrainSet, method="class")
fancyRpartPlot(DecTreeFitModel)

# Test dataset prediction

DecTreePrediction <- predict(DecTreeFitModel, newdata = kbTestSet, type="class")
DecTreeConfMatrix <- confusionMatrix(DecTreePrediction, kbTestSet$classe)

# Plot matrix results

plot(DecTreeConfMatrix$table, col = DecTreeConfMatrix$byClass, main = paste("Decision Tree Accurarcy=", round(DecTreeConfMatrix$overall['Accuracy'], 4)))

```

### iii) Generalized Boosted Model Method

```{r}
# Fit Model
set.seed(12000)
GBMControl <- trainControl(method = "repeatedcv", number = 5, repeats = 1)
GBMModFit <- train(classe ~ ., data=kbTrainSet, method = "gbm", trControl = GBMControl, verbose = FALSE)
GBMModFit$finalModel

# Test Data prediction

GMBPrediction <- predict(GBMModFit, newdata=kbTestSet)
GBMConfMatrix <- confusionMatrix(GMBPrediction, kbTestSet$classe)
GBMConfMatrix

# Matril result plot

plot(GBMConfMatrix$table, col=GBMConfMatrix$byClass, main = paste("GBM - Accuracy", round(GBMConfMatrix$overall['Accuracy'], 4)))

```

# 5. Applying the Selected Model to the Test Data set

### The accuracy of the Three regression methods above are

  i) Radom Forest = 0.9981
 ii) Decision Tree = 0.7346
iii) GBM = 0.99

The Random Forest model will be applied to predict the 20 quiz results as below:
```{r}
TestPrediction <- predict(FitRmodandForest, newdata = kbTesting)
TestPrediction
```

